Toward Resolved Simulations of Burning Fronts in Thermonuclear X-ray Bursts

### 2018-09-30 astronum_2018/astronum-xrb.tex
tweak size
a little more on the SDC

\begin{figure}
\centering
\includegraphics[width=0.95\linewidth]{xrb}
\caption{\label{fig:xrb} Helium flame spreading across the surface of
  a neutron star.  Shown are temperature (top), the mean molecular
  weight of the ash (second), the nuclear energy generation rate
  (third), and the velocity out of the simulation plane (bottom).  The
  out of plane velocity contours indicate where the Coriolis force has
  created a hurricane-like wind structure which confines the flame,
  leading to an inclined front. It is at such an interface that the
  temperature and burning rate are highest.  }
\end{figure}

@@

As an example of the improvements in coupling between hydro
and \MarginPar{Don can add some comments about startup costs?}
reactions, Figure~\ref{fig:sdc} shows the mass fraction of helium
behind a detonation over the course of 3 timesteps.  The points
represent individual calls to the righthand side function of our
reaction network (a 13-isotope $\alpha$-chain), and the density of
points indicates how hard the integrator is working.  We use the same
tolerances for both the Strang-split method and the SDC method.  For
the SDC method, we are using a second-order accurate method based on
\cite{SDC-old}.  We predict a time-centered advection term,
$\Adv{\Uc}^{n+1/2}_i$ using standard unsplit Godunov methods, but
explicitly include a reactive source term in the tracing of the
interface states, ${\bf R}(\Uc_i)$.  We can then solve the reactive
system, using this advective term as a piecewise-constant-in-time
source
\begin{equation}
\label{eq:syst}
\frac{d{\Uc_i}}{dt} = -\Adv{\Uc}^{n+1/2}_i + {\bf R}(\Uc_i)
\end{equation}
This system can be integrated with standard ODE methods.  To achieve
second-order accuracy, we need to iterate, using the updated $\Uc$ to
create a reactive source included in the predictor for the advective
update, and then reintegrate the system.  The result of this iteration
is that the advection sees the effects of reactions over the timestep
and the reactions see the effects of advection.  For the zone tracked
in Figure~\ref{fig:sdc}, each SDC iteration called the righthand side
function of the network 5 times less than the Strang case.
We are continuing to develop this new integration strategy, with full
fourth-order in space and time reacting hydrodynamics methods to
follow shortly.  The XRB simulations are our target application.

\begin{figure}
\centering
\includegraphics[width=0.75\linewidth]{sdc_plot}
\caption{\label{fig:sdc} Helium mass fraction as a function of time
  over three timesteps.  The points represent individual calls to the
  righthand side function of our reaction network (see
  Eq. \ref{eq:syst}) and the density of points indicates how hard the
  integrator is working.}
\end{figure}

>>>>

\begin{figure}
\centering
\includegraphics[width=\linewidth]{xrb}
\caption{\label{fig:xrb} Helium flame spreading across the surface of
  a neutron star.  Shown are temperature (top), the mean molecular
  weight of the ash (second), the nuclear energy generation rate
  (third), and the velocity out of the simulation plane (bottom).  The
  out of plane velocity contours indicate where the Coriolis force has
  created a hurricane-like wind structure which confines the flame,
  leading to an inclined front. It is at such an interface that the
  temperature and burning rate are highest.  }
\end{figure}

@@

As an example of the improvements in coupling between hydro
and \MarginPar{Don can add some comments about startup costs?}
reactions, Figure~\ref{fig:sdc} shows the mass fraction of helium
behind a detonation over the course of 2 timesteps.  For this test,
a 19 isotope reaction network was used.  The points
represent individual calls to the righthand side function of our
reaction network, and the density of
points indicates how hard the integrator is working.  We use the same
tolerances for both the Strang-split method and the SDC method.  For
the SDC method, we are using a second-order accurate method based on
\cite{SDC-old}.  We predict a time-centered advection term,
$\Adv{\Uc}^{n+1/2}_i$ using standard unsplit Godunov methods, but
explicitly include a reactive source term in the tracing of the
interface states, ${\bf R}(\Uc_i)$.  We can then solve the reactive
system, using this advective term as a piecewise-constant-in-time
source
\begin{equation}
\label{eq:syst}
\frac{d{\Uc_i}}{dt} = -\Adv{\Uc}^{n+1/2}_i + {\bf R}(\Uc_i)
\end{equation}
This system can be integrated with standard ODE methods.  To achieve
second-order accuracy, we need to iterate, using the updated $\Uc$ to
create a reactive source included in the predictor for the advective
update, and then reintegrate the system.  The result of this iteration
is that the advection sees the effects of reactions over the timestep
and the reactions see the effects of advection.  For the zone tracked
in Figure~\ref{fig:sdc}, each SDC iteration called the righthand side
function of the network 2 times less than the Strang case.  This
savings is dependent on the network and thermodynamic state, but our
tests thus far show that the SDC methods need fewer network calls to
integrate a timestep.  We also clearly see how the Strang system
evolves in a discontinuous fashion.  We are continuing to develop this
new integration strategy, with full fourth-order in space and time
reacting hydrodynamics methods to follow shortly.  The XRB simulations
are our target application.

\begin{figure}
\centering
\includegraphics[width=0.75\linewidth]{sdc_plot}
\caption{\label{fig:sdc} Helium mass fraction as a function of time
  over two timesteps for a detonation test problem with a 19-isotope
  reaction network.  The points represent individual calls to the
  righthand side function of our reaction network (see
  Eq. \ref{eq:syst}) and the density of points indicates how hard the
  integrator is working.  For the Strang case, we see that the first
  half of the reactions brings us off of the smooth solution
  (represented by the SDC curve), then the advection overcorrects,
  resulting in a discontinuous jump, and finally the second half of
  the reactions works to bring the state back to the solution curve.}
\end{figure}
